---
title: "Manuscript Figures"
output: html_document
---


# Figure 1: Global differences in distributions

```{r}
fig_path <- "/net/irizarryfs01/srv/export/irizarryfs01_backed_up/share_root/shicks/papers/qsmooth"

library(affy)
library(dplyr)
library(quantro)
library(qsmooth)
library(preprocessCore)

# extracted from quantro::matdensity
getdensity <- function(object) {
        min.object <- min(object[is.finite(object)], na.rm = TRUE)
        max.object <- max(object[is.finite(object)], na.rm = TRUE)
        densityMat <- apply(object, 2, function(z) {
            density(z, from = min.object, to = max.object, na.rm = TRUE)$y
        })
        x = seq(from = min.object, to = max.object, length.out = nrow(densityMat))
        list(densityMat = densityMat, x = x)
}


#### smoking data set
library(alveolarSmokingAffyData)
data(alveolarSmokingAffyData)

# extract raw PM values
pmMat <- pm(alveolarSmokingAffyData)
pd <- pData(alveolarSmokingAffyData)

# apply qn and qsmooth
pmMatQN <- preprocessCore::normalize.quantiles(pmMat)
load(file.path(fig_path, "data/pmMatAlveolarSmokingAffyDataQsmooth.RData"))

log2pmMat <- log2(pmMat)
log2pmMatQN <- log2(pmMatQN)
log2pmMatQS <- log2(pmMatAlveolarSmokingAffyDataQsmooth)

lgroup <- c("Nonsmoker (n=15)", "Smoker (n=15)", "Asthmatic (n=15)")

pdf(file.path(fig_path, "Figures/Figure1-supp/Figure1_smokerRaw.pdf"), width = 5, height = 4.5)
matdensity(log2pmMat, groupFactor = pd$Group, xlab = "log2 PM values", 
           ylab = " ", lwd = 2, ylim = c(0, 0.36))
legend('topright', lgroup, lwd = 2, col = 1:3)
dev.off()

pdf(file.path(fig_path, "Figures/Figure1-supp/Figure1_smokerQN.pdf"), width = 5, height = 4.5)
matdensity(log2pmMatQN, groupFactor = pd$Group, xlab = "log2 PM values", 
           ylab = " ", lwd = 2, ylim = c(0, 0.36))
legend('topright', levels(pd$Group), lwd = 2, col = 1:3)
dev.off()

pdf(file.path(fig_path, "Figures/Figure1-supp/Figure1_smokerQS.pdf"), width = 5, height = 4.5)
matdensity(log2pmMatQS, groupFactor = pd$Group, xlab = "log2 PM values", 
           ylab = " ", lwd = 2, ylim = c(0, 0.36))
legend('topright', levels(pd$Group), lwd = 2, col = 1:3)
dev.off()


#### tcruzi data
load(file.path(fig_path, "tcruzi/tcruzi_eset.rda"))
eset <- exprs(tcruzi.eset)
pd <- pData(tcruzi.eset)

eset <- eset[,pd$stage %in% c("epi", "trypo")]
pd <- pd[pd$stage %in% c("epi", "trypo"),]

keep = rowSums(eset) >  ncol(eset)
eset = eset[keep,]
dim(eset)

# apply qn and qsmooth
pmMatQN <- preprocessCore::normalize.quantiles(eset)
pmMatQS <- qsmoothData(qsmooth(eset, groupFactor = pd$stage))

log2pmMat <- log2(eset)
log2pmMatQN <- log2(pmMatQN)
log2pmMatQS <- log2(pmMatQS)

pdf(file.path(fig_path,"Figures/Figure1-supp/Figure1_tcruziRaw.pdf"), width = 5, height = 4.5)
matdensity(log2pmMat, groupFactor = pd$stage, xlab = "log2(count + 1)", 
           ylab = " ", lwd = 2, ylim = c(0, 0.42))
legend('topright', c("T. cruzi epimastigotes (n=3)", "Extracellular trypomastigotes (n=4)"), 
      lwd = 2, col = 1:2)
dev.off()

pdf(file.path(fig_path,"Figures/Figure1-supp/Figure1_tcruziQN.pdf"), width = 5, height = 4.5)
matdensity(log2pmMatQN, groupFactor = pd$stage, xlab = "log2(count + 1)", 
           ylab = " ", lwd = 2, ylim = c(0, 0.42))
legend('topright', c("T. cruzi epimastigotes", "Extracellular trypomastigotes"), 
      lwd = 2, col = 1:2)
dev.off()

pdf(file.path(fig_path,"Figures/Figure1-supp/Figure1_tcruziQS.pdf"), width = 5, height = 4.5)
matdensity(log2pmMatQS, groupFactor = pd$stage, xlab = "log2(count + 1)", 
           ylab = " ", lwd = 2, ylim = c(0, 0.42))
legend('topright', c("T. cruzi epimastigotes", "Extracellular trypomastigotes"), 
      lwd = 2, col = 1:2)
dev.off()


#### brain and liver data set
barcode_path <- "/net/irizarryfs01/srv/export/irizarryfs01_backed_up/share_root/ypitajuarez/Barcode/"

pd <- readRDS(paste0(barcode_path, "/U133plus20/Annotation/Barcode3.tissue.RDS"))
pdSub <- pd %>% filter(tissue %in% c("liver", "brain"), 
                       series %in% c("GSE17612", "GSE21935", "GSE14668", "GSE29721", "GSE6764"))
pdSub$odyFilename <- paste0(paste("U133plus20", pdSub$tissue, pdSub$series, 
                                  pdSub$sample, sep="/"), ".CEL.gz")   
pdSub$odyFilename[which(pdSub$series == "GSE21935")] <- 
  paste("U133plus20", pdSub$tissue[which(pdSub$series == "GSE21935")], 
        pdSub$series[which(pdSub$series == "GSE21935")], 
        list.files(paste0(barcode_path, "U133plus20/brain/GSE21935")), sep="/")
affyDat = ReadAffy(filenames = paste0(barcode_path, pdSub$odyFilename)) # Creates an AffyBatch object
pmMat = pm(affyDat) # raw

# apply qn and qsmooth
pmMatQN <- preprocessCore::normalize.quantiles(pmMat)
load(file.path(fig_path,"data/pmMatBrainLiverAffyDataQsmooth.RData"))

log2pmMat <- log2(pmMat)
log2pmMatQN <- log2(pmMatQN)
log2pmMatQS <- log2(pmMatBrainLiverAffyDataQsmooth)


library(RColorBrewer)
colGreen = brewer.pal(n=7, "Greens")
colRed = brewer.pal(n=7, "Reds")
colCombined = c(colRed[c(5)], colGreen[c(7,5)], colRed[c(7,4)])

# reorder samples
object <- log2pmMat[,c(73:82,11:18,42:53,1:10,19:41,54:72)]
output <- getdensity(object)

pdf(file.path(fig_path, "Figures/Figure1-supp/Figure1_brainLiverRaw.pdf"), width = 5, height = 4.5)
matplot(x = output$x, output$densityMat, type = "l", lty = 1, 
        col = colCombined[factor(pdSub$series[c(73:82,11:18,42:53,1:10,19:41,54:72)])],
        xlab = "log2 PM values", ylab = " ", lwd = 2)
legend('topright', 
       c("Brain (GSE17612, n=23)", "Brain (GSE21935, n=19)", 
         "Liver (GSE29721, n=10)", "Liver (GSE14668, n=20)", "Liver (GSE6764, n=10)"),
       lwd = 2, col = c(colGreen[c(7,5)], colRed[c(7,5,4)]))
dev.off()

object <- log2pmMatQN
output <- getdensity(object)
pdf(file.path(fig_path, "Figures/Figure1-supp/Figure1_brainLiverQN.pdf"), width = 5, height = 4.5)
matplot(x = output$x, output$densityMat, type = "l", lty = 1, 
        col = c(colGreen[6], colRed[6])[factor(pdSub$tissue)], 
        xlab = "log2 PM values", ylab = " ", lwd = 2, ylim = c(0, 1.2))
legend('topright', c("Brain", "Liver"), lwd = 2, col = c(colGreen[6], colRed[6]))
dev.off()

object <- log2pmMatQS
output <- getdensity(object)
pdf(file.path(fig_path, "Figures/Figure1-supp/Figure1_brainLiverQS.pdf"), width = 5, height = 4.5)
matplot(x = output$x, output$densityMat, type = "l", lty = 1, 
        col = c(colGreen[6], colRed[6])[factor(pdSub$tissue)], 
        xlab = "log2 PM values", ylab = " ", lwd = 2, ylim = c(0, 1.2))
legend('topright', c("Brain", "Liver"), lwd = 2, col = c(colGreen[6], colRed[6]))
dev.off()

```


# Figure 2: Density plots using bodymapRat data

```{r}
library(Biobase)
library(bodymapRat) # load data set
library(preprocessCore) # quantile normalize
library(qsmooth) 
library(quantro) # test for global differences in distributions of spike-ins

fig_path <- "/net/irizarryfs01/srv/export/irizarryfs01_backed_up/share_root/shicks/papers/qsmooth/"

data(bodymapRat)
pd = pData(bodymapRat)

# Select samples and filter genes
sel = (pd$organ %in% c("Liver", "Brain")) & 
      (pd$stage %in% c(21)) & (pd$sex %in% c("F", "M"))
keep = rowSums(exprs(bodymapRat[, sel])) > 75
subdataRat = bodymapRat[keep, sel]
dim(subdataRat)

## Density plot
pdf(file.path(fig_path, "Figures/rat_density_liver_brain.pdf"), 
    width = 8, height = 10)
par(mfrow=c(3,2))
par(mar=c(2, 2, 1, .5), mgp=c(0.7, 0.1, 0), tck=0.02)

lpcounts = log2(exprs(subdataRat) + 1)  
matdensity(lpcounts, groupFactor=pData(subdataRat)$organ, 
           ylim = c(0, 0.17), xlab="log2(count + 1)", ylab=" ")

# spike-ins only
lpcounts.raw.ercc = lpcounts[grep("^ERCC", rownames(lpcounts)), ]
matdensity(lpcounts.raw.ercc, groupFactor=pData(subdataRat)$organ, 
           ylim = c(0, 0.13), xlab="log2(count + 1)", ylab=" ")

lpcounts = preprocessCore::normalize.quantiles(log2(exprs(subdataRat)+1))
matdensity(lpcounts, groupFactor=pData(subdataRat)$organ, 
           ylim = c(0, 0.17), xlab="log2(count + 1)", ylab=" ")

# spike-ins only
lpcounts.qn.ercc = lpcounts[grep("^ERCC", rownames(lpcounts)), ]
matdensity(lpcounts.qn.ercc, groupFactor=pData(subdataRat)$organ, 
           ylim = c(0, 0.13), xlab="log2(count + 1)", ylab=" ")

counts = exprs(subdataRat)
lpcounts = qsmoothData(qsmooth(log2(counts+1), groupFactor = pData(subdataRat)$organ))
matdensity(lpcounts, groupFactor=pData(subdataRat)$organ, 
           ylim = c(0, 0.17), xlab="log2(count + 1)", ylab=" ")

# spike-ins only
lpcounts.qs.ercc = lpcounts[grep("^ERCC", rownames(lpcounts)), ]
matdensity(lpcounts.qs.ercc, groupFactor=pData(subdataRat)$organ, 
           ylim = c(0, 0.13), xlab="log2(count + 1)", ylab=" ")
dev.off()
```

#### Test for differences in distribution using KS test

```{r}
ks.test(c(lpcounts.raw.ercc[,pData(subdataRat)$organ %in% "Brain"]), 
        c(lpcounts.raw.ercc[,pData(subdataRat)$organ %in% "Liver"]))
ks.test(c(lpcounts.qn.ercc[,pData(subdataRat)$organ %in% "Brain"]), 
        c(lpcounts.qn.ercc[,pData(subdataRat)$organ %in% "Liver"]))
ks.test(c(lpcounts.qs.ercc[,pData(subdataRat)$organ %in% "Brain"]), 
        c(lpcounts.qs.ercc[,pData(subdataRat)$organ %in% "Liver"]))
```


# Figure 3: Density plots using GTEx data

Scripts for Figure 3 are [available on Dropbox here](https://www.dropbox.com/sh/qs3dx36bbvcpufm/AADooxO7x0A89T6gGeH6oXvGa?dl=0). 


# Figure 4: Gene-specific effects using GTEx data

Scripts for Figure 4 are [available on Dropbox here](https://www.dropbox.com/sh/qs3dx36bbvcpufm/AADooxO7x0A89T6gGeH6oXvGa?dl=0). 

# Figure 5: Density plots, boxplots and PCA plots using Flowsorted DNAm data

```{r}
fig_path <- "/net/irizarryfs01/srv/export/irizarryfs01_backed_up/share_root/shicks/papers/qsmooth/"

library(FlowSorted.Blood.450k) # load data set
library(quantro)
library(qsmooth)

# extract raw beta values
cellcompMethyl <- preprocessRaw(updateObject(FlowSorted.Blood.450k))
p <- getBeta(cellcompMethyl, type = "Illumina")
pd = as.data.frame(pData(cellcompMethyl))

IDs = c("CD8T","CD4T", "NK","Bcell","Mono","Gran")
p0 = p[,which(pd$CellType %in% IDs)]
pd0 = pd[which(pd$CellType %in% IDs),]

# remove outlier DNAm sample
p0 = p0[,-which(pd0$Sample_Name == "CD8+_105")]
pd0 = pd0[-which(pd0$Sample_Name == "CD8+_105"),]

# create figure
pdf(file.path(fig_path, "Figures/FlowSorted_density_boxplot_pca.pdf"),
    width = 12, height = 12)

# raw 
par(mfrow=c(3,3))
matdensity(p0, groupFactor = pd0$CellType, xlab = "beta values", ylab = "density", 
           main = "raw (no normalization)")
legend('top', levels(factor(pd0$CellType)), lty = 1, col = 1:6, lwd = 3)

matboxplot(p0, groupFactor = pd0$CellType, main = "raw (no normalization)", 
          ylab = "beta values", names = pd0$CellType)

s.raw <- svd(p0)
scores <- data.frame(pd0, s.raw$v[, 1:4])
plot(scores$X1, scores$X2, pch = 16, col = c(1:6)[factor(pd0$CellType)], 
     xlab = "PC1", ylab = "PC2",  main = "raw (no normalization)")

# qn
p0.qn <- preprocessCore::normalize.quantiles(p0)
matdensity(p0.qn, groupFactor = pd0$CellType,  main = "quantile normalization", 
           xlab = "beta values", ylab = "density")
legend('top', levels(factor(pd0$CellType)), lty = 1, col = 1:6, lwd = 3)
matboxplot(p0.qn, groupFactor = pd0$CellType, main = "quantile normalization", 
          ylab = "beta values", names = pd0$CellType)
s.qn <- svd(p0.qn)
scores <- data.frame(pd0, s.qn$v[, 1:4])
plot(scores$X1, scores$X2, pch = 16, col =  c(1:6)[factor(pd0$CellType)], 
     xlab = "PC1", ylab = "PC2",  main = "quantile normalization")

# qsmooth
p0.qs <- qsmoothData(qsmooth(p0, groupFactor = factor(pd0$CellType)))
matdensity(p0.qs, groupFactor = pd0$CellType, main = "qsmooth", 
          xlab = "beta values", ylab = "density")
legend('top', levels(factor(pd0$CellType)), lty = 1, col = 1:6, lwd = 3)
matboxplot(p0.qs, groupFactor = pd0$CellType, main = "qsmooth", 
          ylab = "beta values", names = pd0$CellType)
s.qs <- svd(p0.qs)
scores <- data.frame(pd0, s.qs$v[, 1:4])
plot(scores$X1, scores$X2, pch = 16, col =  c(1:6)[factor(pd0$CellType)], 
     xlab = "PC1", ylab = "PC2",  main = "qsmooth")
dev.off()

```

