---
title: "Supplemental Figures"
output: html_document
---


# Supplemental Figure 1: Boxplots using bodymapRat data using raw, QN, qsmooth

```{r}
library(Biobase)
library(bodymapRat) # load data set
library(preprocessCore) # quantile normalize
library(quantro) # test for global differences in distributions of spike-ins

fig_path <- "/net/irizarryfs01/srv/export/irizarryfs01_backed_up/share_root/shicks/papers/qsmooth/"

library(bodymapRat)
data(bodymapRat)
pd = pData(bodymapRat)

# Select samples and filter genes
sel = (pd$organ %in% c("Liver", "Brain")) & 
      (pd$stage %in% c(21)) & (pd$sex %in% c("F", "M"))
keep = rowSums(exprs(bodymapRat[, sel])) > 75
subdataRat = bodymapRat[keep, sel]
dim(subdataRat)

## Density plot
pdf(file.path(fig_path, "Figures/rat_boxplot_liver_brain1.pdf"), 
    width = 8, height = 10)
par(mfrow=c(3,2))
par(mar=c(2, 2, 1, .5), mgp=c(0.7, 0.1, 0), tck=0.02)

# raw (no normalization)
lpcounts = log2(exprs(subdataRat) + 1)  
matboxplot(lpcounts, groupFactor=pData(subdataRat)$organ, 
           ylab="log2(count + 1)", xlab=" ")

lpcounts.raw.ercc = lpcounts[grep("^ERCC", rownames(lpcounts)), ]
matboxplot(lpcounts.raw.ercc, groupFactor=pData(subdataRat)$organ, 
           ylab="log2(count + 1)", xlab=" ")

# quantile normalization
lpcounts = preprocessCore::normalize.quantiles(log2(exprs(subdataRat)+1))
matboxplot(lpcounts, groupFactor=pData(subdataRat)$organ, 
           ylab="log2(count + 1)", xlab=" ")

lpcounts.qn.ercc = lpcounts[grep("^ERCC", rownames(lpcounts)), ]
matboxplot(lpcounts.qn.ercc, groupFactor=pData(subdataRat)$organ, 
           ylab="log2(count + 1)", xlab=" ")

# qsmooth
counts = exprs(subdataRat)
lpcounts = qsmoothData(qsmooth(log2(counts+1), groupFactor = pData(subdataRat)$organ))
matboxplot(lpcounts, groupFactor=pData(subdataRat)$organ, 
           ylab="log2(count + 1)", xlab=" ")

# spike-ins only
lpcounts.qs.ercc = lpcounts[grep("^ERCC", rownames(lpcounts)), ]
matboxplot(lpcounts.qs.ercc, groupFactor=pData(subdataRat)$organ, 
           ylab="log2(count + 1)", xlab=" ")
dev.off()

```



# Supplemental Figure 2: Boxplots using bodymapRat data using RLE, median, ERCC

```{r}
library(Biobase)
library(bodymapRat) # load data set
library(preprocessCore) # quantile normalize
library(qsmooth) 
library(quantro) # test for global differences in distributions of spike-ins

fig_path <- "/net/irizarryfs01/srv/export/irizarryfs01_backed_up/share_root/shicks/papers/qsmooth/"

library(bodymapRat)
data(bodymapRat)
pd = pData(bodymapRat)

# Select samples and filter genes
sel = (pd$organ %in% c("Liver", "Brain")) & 
      (pd$stage %in% c(21)) & (pd$sex %in% c("F", "M"))
keep = rowSums(exprs(bodymapRat[, sel])) > 75
subdataRat = bodymapRat[keep, sel]
dim(subdataRat)

## Density plot
pdf(file.path(fig_path, "Figures/rat_boxplot_liver_brain2.pdf"), 
    width = 8, height = 10)
par(mfrow=c(3,2))
par(mar=c(2, 2, 1, .5), mgp=c(0.7, 0.1, 0), tck=0.02)

# DESeq2 RLE normalization 
lpcounts = log2(exprs(subdataRat) + 1)
scal = apply(lpcounts - rowMeans(lpcounts), 2, median)
lpcounts = t(t(lpcounts) - scal)
matboxplot(lpcounts, groupFactor=pData(subdataRat)$organ, 
           ylab="log2(count + 1)", xlab=" ")

lpcounts.raw.ercc = lpcounts[grep("^ERCC", rownames(lpcounts)), ]
matboxplot(lpcounts.raw.ercc, groupFactor=pData(subdataRat)$organ, 
           ylab="log2(count + 1)", xlab=" ")

# median normalization
lpcounts = log2(exprs(subdataRat) + 1)
lpcounts = t(t(lpcounts) - apply(lpcounts, 2, median))
matboxplot(lpcounts, groupFactor=pData(subdataRat)$organ, 
           ylab="log2(count + 1)", xlab=" ")

lpcounts.qn.ercc = lpcounts[grep("^ERCC", rownames(lpcounts)), ]
matboxplot(lpcounts.qn.ercc, groupFactor=pData(subdataRat)$organ, 
           ylab="log2(count + 1)", xlab=" ")

# scale by median of spike-ins
counts = exprs(subdataRat)
pcounts = counts + 1
med.ercc = apply(pcounts[grep("^ERCC", rownames(pcounts)), ], 2, median)
ercc.scaled.counts = t(t(pcounts) / med.ercc)
ercc.scaled.counts = ercc.scaled.counts / min(ercc.scaled.counts)
lpcounts = log2(ercc.scaled.counts)
matboxplot(lpcounts, groupFactor=pData(subdataRat)$organ, 
           ylab="log2(count + 1)", xlab=" ")

lpcounts.qs.ercc = lpcounts[grep("^ERCC", rownames(lpcounts)), ]
matboxplot(lpcounts.qs.ercc, groupFactor=pData(subdataRat)$organ, 
           ylab="log2(count + 1)", xlab=" ")
dev.off()

```


# Supplemental Figure 3: Density plot of GTEx data

Scripts for Supplemental Figure 3 are [available on Dropbox here](https://www.dropbox.com/sh/qs3dx36bbvcpufm/AADooxO7x0A89T6gGeH6oXvGa?dl=0). 


# Supplemental Figure 4: Density plots, boxplots and PCA plots using Flowsorted DNAm data (M values)

```{r}
fig_path <- "/net/irizarryfs01/srv/export/irizarryfs01_backed_up/share_root/shicks/papers/qsmooth/"

library(FlowSorted.Blood.450k) # load data set
library(quantro)
library(qsmooth)

# extract raw beta values
cellcompMethyl <- preprocessRaw(updateObject(FlowSorted.Blood.450k))
M <- getM(cellcompMethyl)
pd = as.data.frame(pData(cellcompMethyl))

IDs = c("CD8T","CD4T", "NK","Bcell","Mono","Gran")
M0 = M[,which(pd$CellType %in% IDs)]
pd0 = pd[which(pd$CellType %in% IDs),]

# remove outlier DNAm sample
M0 = M0[,-which(pd0$Sample_Name == "CD8+_105")]
pd0 = pd0[-which(pd0$Sample_Name == "CD8+_105"),]

tmp = apply(M0, 1, function(x){ any(x %in% c(-Inf, Inf)) })
M0 <- na.omit(M0[!tmp, ])


# create figure
pdf(file.path(fig_path, "Figures/FlowSorted_density_boxplot_pca_Mvalues.pdf"),
    width = 12, height = 12)

# raw 
par(mfrow=c(3,3))
matdensity(M0, groupFactor = pd0$CellType, xlab = "M values", ylab = "density", 
           main = "raw (no normalization)", xlim = c(-7, 7))
legend('top', levels(factor(pd0$CellType)), lty = 1, col = 1:6, lwd = 3)

matboxplot(M0, groupFactor = pd0$CellType, main = "raw (no normalization)", 
          ylab = "M values", names = pd0$CellType)

s.raw <- svd(M0)
scores <- data.frame(pd0, s.raw$v[, 1:4])
plot(scores$X1, scores$X2, pch = 16, col = c(1:6)[factor(pd0$CellType)], 
     xlab = "PC1", ylab = "PC2",  main = "raw (no normalization)")

# qn
M0.qn <- preprocessCore::normalize.quantiles(M0)
matdensity(M0.qn, groupFactor = pd0$CellType,  main = "quantile normalization", 
           xlab = "M values", ylab = "density", xlim=c(-7, 7))
legend('top', levels(factor(pd0$CellType)), lty = 1, col = 1:6, lwd = 3)
matboxplot(M0.qn, groupFactor = pd0$CellType, main = "quantile normalization", 
          ylab = "M values", names = pd0$CellType)
s.qn <- svd(M0.qn)
scores <- data.frame(pd0, s.qn$v[, 1:4])
plot(scores$X1, scores$X2, pch = 16, col =  c(1:6)[factor(pd0$CellType)], 
     xlab = "PC1", ylab = "PC2",  main = "quantile normalization")

# qsmooth
M0.qs <- qsmoothData(qsmooth(M0, groupFactor = factor(pd0$CellType)))
matdensity(M0.qs, groupFactor = pd0$CellType, main = "qsmooth", 
          xlab = "M values", ylab = "density", xlim=c(-7, 7))
legend('top', levels(factor(pd0$CellType)), lty = 1, col = 1:6, lwd = 3)
matboxplot(M0.qs, groupFactor = pd0$CellType, main = "qsmooth", 
          ylab = "M values", names = pd0$CellType)
s.qs <- svd(M0.qs)
scores <- data.frame(pd0, s.qs$v[, 1:4])
plot(scores$X1, scores$X2, pch = 16, col =  c(1:6)[factor(pd0$CellType)], 
     xlab = "PC1", ylab = "PC2",  main = "qsmooth")
dev.off()

```


# Supplemental Figure 5: Bias variance tradeoff 

See the `biasVarianceTradeoff.Rmd` in the [qsmoothPaper/simulationStudies repository](https://github.com/stephaniehicks/qsmoothPaper/blob/master/scripts/simulationStudies/biasVarianceTradeoff.Rmd)


# Supplemental Figure 6: normal and cancer bladder samples

```{r}
library(quantro)
library(qsmooth)
library(sva)
library(bladderbatch)
data(bladderdata)
pd <- pData(bladderEset)

bladderEset <- bladderEset[,pd$cancer %in% c("Cancer", "Normal")]
pd <- pData(bladderEset)
pd$cancer <- factor(pd$cancer)
pd$batch <- factor(pd$batch)
eset <- exprs(bladderEset)

mod = model.matrix(~cancer, data=pd)
mod0 = model.matrix(~1,data=pd)
modBatch = model.matrix(~cancer + batch ,data=pd)
mod0Batch = model.matrix(~batch,data=pd)

table(pd$cancer, pd$batch)

pdf(file.path(fig_path, "Figures/bladderbatch_density.pdf"),
    width = 12, height = 8)
par(mfrow=c(2,3))
# raw (colors by cancer and by batch)
matdensity(eset, pd$cancer, main = "No normalization\n(colored by cancer)", 
           xlab = "Gene expression", ylab = "Density", ylim=c(0, 0.3))
legend('topright', c("Cancer (n=40)", "Normal (n=8)"), lty = 1, col = 1:3, lwd = 3)
matdensity(eset, pd$batch, main = "No normalization\n(colored by batch)", 
           xlab = "Gene expression", ylab = "Density", ylim=c(0, 0.3))
legend('topright', c("Batch 1 (n=11)", "Batch 2 (n=18)", "Batch 3 (n=4)", "Batch 5 (n=15)"), 
       lty = 1, col = 1:5, lwd = 3)
pValuesRaw = f.pvalue(eset,mod,mod0)
qValuesRaw = p.adjust(pValuesRaw,method="BH")
mean(qValuesRaw < 0.05)

# combat
combat_edata = ComBat(dat=eset, batch=pd$batch, mod=mod0)
matdensity(combat_edata, pd$cancer, main = "Adjusted for batches using ComBat", 
           xlab = "Gene expression", ylab = "Density", ylim=c(0, 0.3))
legend('topright', c("Cancer (n=40)", "Normal (n=8)"), lty = 1, col = 1:3, lwd = 3)
pValuesComBat = f.pvalue(combat_edata,mod,mod0)
qValuesComBat = p.adjust(pValuesComBat,method="BH")
mean(qValuesComBat < 0.05)

# quantile normalize
qn <- preprocessCore::normalize.quantiles(eset)
matdensity(qn, pd$cancer, main = "Quantile Normalization", 
           xlab = "Gene expression", ylab = "Density", ylim=c(0, 0.3))
legend('topright', c("Cancer (n=40)", "Normal (n=8)"), lty = 1, col = 1:3, lwd = 3)

pValuesQN = f.pvalue(qn,mod,mod0)
qValuesQN = p.adjust(pValuesQN,method="BH")
mean(qValuesQN < 0.05)

# qsmooth with cancer status
qsNorm <- qsmooth(eset, pd$cancer)
matdensity(qsmoothData(qsNorm), pd$cancer, main = "qsmooth", 
           xlab = "Gene expression", ylab = "Density", ylim=c(0, 0.3))
legend('topright', c("Cancer (n=40)", "Normal (n=8)"), lty = 1, col = 1:3, lwd = 3)
pValuesqsNorm = f.pvalue(qsmoothData(qsNorm), mod, mod0)
qValuesqsNorm = p.adjust(pValuesqsNorm,method="BH")
mean(qValuesqsNorm < 0.05)

# qsmooth with cancer status, removing batch
qsNorm.withoutBatch <- qsmooth(eset, pd$cancer, batch = pd$batch) # remove batch
matdensity(qsmoothData(qsNorm.withoutBatch), pd$cancer, 
           main = "qsmooth\n(After adjusting for batches using ComBat)",
           xlab = "Gene expression", ylab = "Density", ylim=c(0, 0.3))
legend('topright', c("Cancer (n=40)", "Normal (n=8)"), lty = 1, col = 1:3, lwd = 3)
pValuesqsNorm.withoutBatch = f.pvalue(qsmoothData(qsNorm.withoutBatch), mod, mod0)
qValuesqsNorm.withoutBatch = p.adjust(pValuesqsNorm.withoutBatch,method="BH")
mean(qValuesqsNorm.withoutBatch < 0.05)

dev.off()
```

