---
title: "Supplemental Figures"
output: html_document
---


# Supplemental Figure 1: Boxplots using bodymapRat data using raw, QN, qsmooth

```{r}
library(Biobase)
library(bodymapRat) # load data set
library(preprocessCore) # quantile normalize
library(quantro) # test for global differences in distributions of spike-ins

fig_path <- "/net/irizarryfs01/srv/export/irizarryfs01_backed_up/share_root/shicks/papers/qsmooth/"

library(bodymapRat)
data(bodymapRat)
pd = pData(bodymapRat)

# Select samples and filter genes
sel = (pd$organ %in% c("Liver", "Brain")) & 
      (pd$stage %in% c(21)) & (pd$sex %in% c("F", "M"))
keep = rowSums(exprs(bodymapRat[, sel])) > 75
subdataRat = bodymapRat[keep, sel]
dim(subdataRat)

## Density plot
pdf(file.path(fig_path, "Figures/rat_boxplot_liver_brain1.pdf"), 
    width = 8, height = 10)
par(mfrow=c(3,2))
par(mar=c(2, 2, 1, .5), mgp=c(0.7, 0.1, 0), tck=0.02)

# raw (no normalization)
lpcounts = log2(exprs(subdataRat) + 1)  
matboxplot(lpcounts, groupFactor=pData(subdataRat)$organ, 
           ylab="log2(count + 1)", xlab=" ")

lpcounts.raw.ercc = lpcounts[grep("^ERCC", rownames(lpcounts)), ]
matboxplot(lpcounts.raw.ercc, groupFactor=pData(subdataRat)$organ, 
           ylab="log2(count + 1)", xlab=" ")

# quantile normalization
lpcounts = preprocessCore::normalize.quantiles(log2(exprs(subdataRat)+1))
matboxplot(lpcounts, groupFactor=pData(subdataRat)$organ, 
           ylab="log2(count + 1)", xlab=" ")

lpcounts.qn.ercc = lpcounts[grep("^ERCC", rownames(lpcounts)), ]
matboxplot(lpcounts.qn.ercc, groupFactor=pData(subdataRat)$organ, 
           ylab="log2(count + 1)", xlab=" ")

# qsmooth
counts = exprs(subdataRat)
lpcounts = qsmoothData(qsmooth(log2(counts+1), groupFactor = pData(subdataRat)$organ))
matboxplot(lpcounts, groupFactor=pData(subdataRat)$organ, 
           ylab="log2(count + 1)", xlab=" ")

# spike-ins only
lpcounts.qs.ercc = lpcounts[grep("^ERCC", rownames(lpcounts)), ]
matboxplot(lpcounts.qs.ercc, groupFactor=pData(subdataRat)$organ, 
           ylab="log2(count + 1)", xlab=" ")
dev.off()

```



# Supplemental Figure 2: Boxplots using bodymapRat data using RLE, median, ERCC

```{r}
library(Biobase)
library(bodymapRat) # load data set
library(preprocessCore) # quantile normalize
library(qsmooth) 
library(quantro) # test for global differences in distributions of spike-ins

fig_path <- "/net/irizarryfs01/srv/export/irizarryfs01_backed_up/share_root/shicks/papers/qsmooth/"

library(bodymapRat)
data(bodymapRat)
pd = pData(bodymapRat)

# Select samples and filter genes
sel = (pd$organ %in% c("Liver", "Brain")) & 
      (pd$stage %in% c(21)) & (pd$sex %in% c("F", "M"))
keep = rowSums(exprs(bodymapRat[, sel])) > 75
subdataRat = bodymapRat[keep, sel]
dim(subdataRat)

## Density plot
pdf(file.path(fig_path, "Figures/rat_boxplot_liver_brain2.pdf"), 
    width = 8, height = 10)
par(mfrow=c(3,2))
par(mar=c(2, 2, 1, .5), mgp=c(0.7, 0.1, 0), tck=0.02)

# DESeq2 RLE normalization 
lpcounts = log2(exprs(subdataRat) + 1)
scal = apply(lpcounts - rowMeans(lpcounts), 2, median)
lpcounts = t(t(lpcounts) - scal)
matboxplot(lpcounts, groupFactor=pData(subdataRat)$organ, 
           ylab="log2(count + 1)", xlab=" ")

lpcounts.raw.ercc = lpcounts[grep("^ERCC", rownames(lpcounts)), ]
matboxplot(lpcounts.raw.ercc, groupFactor=pData(subdataRat)$organ, 
           ylab="log2(count + 1)", xlab=" ")

# median normalization
lpcounts = log2(exprs(subdataRat) + 1)
lpcounts = t(t(lpcounts) - apply(lpcounts, 2, median))
matboxplot(lpcounts, groupFactor=pData(subdataRat)$organ, 
           ylab="log2(count + 1)", xlab=" ")

lpcounts.qn.ercc = lpcounts[grep("^ERCC", rownames(lpcounts)), ]
matboxplot(lpcounts.qn.ercc, groupFactor=pData(subdataRat)$organ, 
           ylab="log2(count + 1)", xlab=" ")

# scale by median of spike-ins
counts = exprs(subdataRat)
pcounts = counts + 1
med.ercc = apply(pcounts[grep("^ERCC", rownames(pcounts)), ], 2, median)
ercc.scaled.counts = t(t(pcounts) / med.ercc)
ercc.scaled.counts = ercc.scaled.counts / min(ercc.scaled.counts)
lpcounts = log2(ercc.scaled.counts)
matboxplot(lpcounts, groupFactor=pData(subdataRat)$organ, 
           ylab="log2(count + 1)", xlab=" ")

lpcounts.qs.ercc = lpcounts[grep("^ERCC", rownames(lpcounts)), ]
matboxplot(lpcounts.qs.ercc, groupFactor=pData(subdataRat)$organ, 
           ylab="log2(count + 1)", xlab=" ")
dev.off()

```


# Supplemental Figure 3: Density plot of GTEx data

Scripts for Supplemental Figure 3 are [available on Dropbox here](https://www.dropbox.com/sh/qs3dx36bbvcpufm/AADooxO7x0A89T6gGeH6oXvGa?dl=0). 


# Supplemental Figure 4: RMSE comparing QN and qsmooth on GTEx data

Scripts for Supplemental Figure 4 are [available on Dropbox here](https://www.dropbox.com/sh/qs3dx36bbvcpufm/AADooxO7x0A89T6gGeH6oXvGa?dl=0). 


# Supplemental Figure 5: Bias variance tradeoff 

See the `biasVarianceTradeoff.Rmd` in the [qsmoothPaper/simulationStudies repository](https://github.com/stephaniehicks/qsmoothPaper/blob/master/scripts/simulationStudies/biasVarianceTradeoff.Rmd)

